---
title: "SuperGauss"
author: "Martin Lysy, Yun Ling"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SuperGauss}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
## Package Installation

## Simulating a 2-d fBM time series
Simulation is done using function `rSnorm`. You can only simulate stationary Gaussian process with this function. For those unstationary processes but have stationary increments, you can first simulate their increments and then recover the target time series using function `cumsum`. We already prepare the acf function for some common stationary Gaussain process and stationary increment process, like the fBm process increment, exponential decay process and matern process.
```{r}
require(SuperGauss)
N <- 2000 # length of time series
dT <- 1/60 # frequency of observation
alpha <- 0.8 # fBM parameter
acf <- fbm.acf(alpha = alpha, dT = dT, N = N) # obtain the autocorrelation function of the fBM increments
dX <- rSnorm(n = 2, acf = acf) # simulating the fBM increments
Xt <- apply(dX, 2, cumsum) # recovering the fBM process
tSeq <- 1:N * dT
plot(x = tSeq, y = Xt[, 1], ylim = range(Xt), col = "blue", type = "l", xlab = "", ylab = "", 
     main = "simulation of 2-d fBM time series")
par(new = TRUE)
plot(x = tSeq, y = Xt[, 2], ylim = range(Xt), col = "red", type = "l", xlab = "", ylab = "")
```

## Inference

### One parameter case
In this case we are inferencing time series $Z_t = \left[Z_1(t), Z_2(t)\right]$ with each column being i.i.d. fBM process with parameter $\alpha$. Its increment $\Delta Z_t$ follows matrix normal distribution $MN(0, V(\alpha), I)$ where $V(\alpha)$ is the variance matrix of fBM increment, which is a Toeplitz matrix. $I$ is a $2 \times 2$ identity matrix. 

With function `dSnorm`, `Snorm.grad` and `Snorm.Hess` in `SuperGauss` library, we can easily compute the log-likelihood, score function and Hessian matrix of stationary Gaussian system in *superfast* speed $O(n \log^2 n)$. Then we can function `nlm` together with the computed density, score function and Hessian matrix to find the MLE of this stationary Gaussian system.

```{r}
# functions for computing the derivative of the acf of fBM increment
fbm.acf.grad <- function(alpha, dT, N) {
  if(N == 1) {
    dacf <- dT^alpha * log(alpha)
  } else {
    dacf <- c(0, (dT*(1:N))^alpha * log(dT*(1:N)))
    dacf <- .5 * (dacf[1:N+1] + c(dacf[2], dacf[1:(N-1)]) - 2*dacf[1:N])
  }
  dacf
}

# functions for computing the second derivative of the acf of fBM increment
fbm.acf.hess <- function(alpha, dT, N) {
  if(N == 1) {
    d2acf <- dT^alpha * log(alpha)^2
  } else {
    d2acf <- c(0, (dT*(1:N))^alpha * log(dT*(1:N))^2)
    d2acf <- .5 * (d2acf[1:N+1] + c(d2acf[2], d2acf[1:(N-1)]) - 2*d2acf[1:N])
  }
  d2acf
}

# Newton-Raphson Using Function 'nlm'
fbm.loglik <- function(alpha, Xt, dT, acf){
  if(alpha <= 0 || alpha >= 2){
    density <- Inf
  }else{
    dX <- apply(Xt, 2, diff)
    N <- nrow(dX)
    
    acf1 <- fbm.acf(alpha = alpha, dT = dT, N = N)
    acf$setAcf(acf1)
    dacf <- fbm.acf.grad(alpha = alpha, dT = dT, N = N)
    d2acf <- fbm.acf.hess(alpha = alpha, dT = dT, N = N)

    density <- -1 * dSnorm(X = dX, mu = 0, acf = acf, log = TRUE)
    acf$setAcf(acf1)
    attr(density, "gradient") <- -1 * Snorm.grad(X = dX, acf = acf, dacf = dacf)
    acf$setAcf(acf1)
    attr(density, "hessian") <- -1 * Snorm.Hess(X = dX, acf = acf, dacf = dacf, d2acf = d2acf)
  }
  density
}

acf <- Toeplitz(N - 1)
theta.nlm <- nlm(f = fbm.loglik, p = 1, X = Xt, dT = dT, acf = acf)
signif(c(alpha, theta.nlm$estimate), digits = 3)
```

### Two parameter case
In this case, we are inferencing a very commonly-applied time series model **Static Error Model** in microscopy. Here is the model:
$$X_t = \mu t + \left(Z_t + \rho \Psi_t\right)\Sigma$$
where $X_t = \left[X_1(t), X_2(t)\right]$ is the 2-d time series, $\mu = \left[\mu_1, \mu_2\right]$ is the linear drift parameter, $Z_t = \left[Z_1(t), Z_2(t)\right]$ each column is i.i.d. fBM process with parameter $\alpha$, $\rho$ is the weight parameter bewteen $Z_t$ and noise $\Psi$, $\Sigma$ is the 2-d plane variance matrix.

In this demonstration, we assume that $\mu = 0$ and $\Sigma = I$ for simplification. Thus the only two parameters in this model are $\{\alpha, \rho\}$. Here is the inference using function `nlm`:
```{r}
# simulating the matern process
alpha <- 0.8
rho <- 0.2
dT <- 1/30
N <- 2000
acf <- fbm.acf(alpha = alpha, dT = dT, N = N)
dX <- rSnorm(n = 2, acf = acf)
Xt <- apply(dX, 2, cumsum)
Xt <- Xt + rho * matrix(rnorm(2*N), N, 2)
tSeq <- 1:N * dT
plot(x = tSeq, y = Xt[, 1], ylim = range(Xt), col = "blue", type = "l", xlab = "", ylab = "", 
     main = "simulation of 2-d static error time series")
par(new = TRUE)
plot(x = tSeq, y = Xt[, 2], ylim = range(Xt), col = "red", type = "l", xlab = "", ylab = "")


# functions for computing the acf of static error time series increment
static.acf <- function(theta, dT, N){
  alpha <- theta[1]
  rho <- theta[2]
  acf <- fbm.acf(alpha = alpha, dT = dT, N = N) + rho^2 * c(2, -1, rep(0, N - 2))
  acf
}

# functions for computing the derivative of the acf of static error time series increment
static.acf.grad <- function(theta, dT, N) {
  alpha <- theta[1]
  rho <- theta[2]
  dacf <- matrix(NA, N, 2)
  if(N == 1) {
    dacf[, 1] <- dT^alpha * log(alpha)
    dacf[, 2] <- 4 * rho
  } else {
    dacf1 <- c(0, (dT*(1:N))^alpha * log(dT*(1:N)))
    dacf[, 1] <- .5 * (dacf1[1:N+1] + c(dacf1[2], dacf1[1:(N-1)]) - 2*dacf1[1:N])
    dacf[, 2] <- 2 * rho * c(2, -1, rep(0, N-2))
  }
  dacf
}

# functions for computing the secode derivative of the acf of static error time series increment
static.acf.hess <- function(theta, dT, N) {
  alpha <- theta[1]
  rho <- theta[2]
  d2acf <- array(0, c(N, 2, 2))
  if(N == 1) {
    d2acf[, 1, 1] <- dT^alpha * log(alpha)^2
    d2acf[, 2, 1] <- 0
    d2acf[, 1, 2] <- 0
    d2acf[, 2, 2] <- 4
  } else {
    d2acf1 <- c(0, (dT*(1:N))^alpha * log(dT*(1:N))^2)
    d2acf[, 1, 1] <- .5 * (d2acf1[1:N+1] + c(d2acf1[2], d2acf1[1:(N-1)]) - 2*d2acf1[1:N])
    d2acf[, 1, 2] <- rep(0, N)
    d2acf[, 2, 1] <- rep(0, N)
    d2acf[, 2, 2] <- 2 * c(2, -1, rep(0, N-2))
  }
  d2acf
}

# Newton-Raphson Using Function 'nlm'

static.loglik <- function(theta, Xt, dT, acf){
  if(theta[1] <= 0 || theta[1] >= 2){
    density <- Inf
  }else{
    dX <- apply(Xt, 2, diff)
    N <- nrow(dX)
    
    acf1 <- static.acf(theta = theta, dT = dT, N = N)
    dacf <- static.acf.grad(theta = theta, dT = dT, N = N)
    d2acf <- static.acf.hess(theta = theta, dT = dT, N = N)
    acf$setAcf(acf1)
    density <- -1 * dSnorm(X = dX, acf = acf, log = TRUE)
    attr(density, "gradient") <- -1 * Snorm.grad(X = dX, acf = acf, dacf = dacf)
    acf$setAcf(acf1)
    attr(density, "hessian") <- -1 * Snorm.Hess(X = dX, acf = acf, dacf = dacf, d2acf = d2acf)
  }
  density
}

acf <- Toeplitz(N - 1)
theta.nlm <- nlm(f = static.loglik, p = c(1, 0.1), X = Xt, dT = dT, acf = acf)
signif(c(alpha, rho, theta.nlm$estimate), digits = 3)
```